---
title: "chapter6.Rmd"
author: "Emmi-Leena Ihantola"
date: "9 12 2019"
output: html_document
---

# Excercise 6. Analysis of longitudinal data

## Analysis of RATS dataset


### Reading and exporing data
```{r}
library(dplyr)
library(tidyr)
RATSL <-read.csv("RATSL.csv")
glimpse(RATSL)
head(RATSL); tail(RATSL)
```

### Figure 1
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

p1 <- ggplot(RATSL, aes(x = Time, y = Weight, group = ID))
p2 <- p1 + geom_text(aes(label = Group))
p3 <- p2 + scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10))
p4 <- p3 + scale_y_continuous(name = "Weight (grams)")
p5 <- p4 + theme_bw()
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6
```

### Standardization and Figure 2

```{r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()
glimpse(RATSL)


p1 <- ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(name = "standardized weight")
p6
```

### Summary of the dataset and Figure 3
```{r}
n <- RATSL$Time %>% unique() %>% length()
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean=mean(Weight), se=sd(Weight)/sqrt(n) ) %>%
  ungroup()
glimpse(RATSS)
p1 <- ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group))
p2 <- p1 + geom_line() + scale_linetype_manual(values = c(1,2,3))
p3 <- p2 + geom_point(size=3) + scale_shape_manual(values = c(1,2,3))
p4 <- p3 + geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3)
p5 <- p4 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6 <- p5 + theme(legend.position = c(0.8,0.8))
p7 <- p6 + scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
p7
```
### Figure 4
```{r}
p1 <- ggplot(RATSL, aes(x = factor(Time), y = Weight, fill = Group))
p2 <- p1 + geom_boxplot(position = position_dodge(width = 0.9))
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + theme(legend.position = c(0.8,0.8))
p5 <- p4 + scale_x_discrete(name = "Time")
p5
```


### Figure 5
```{r}
RATSL8S <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
glimpse(RATSL8S)
p1 <- ggplot(RATSL8S, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight), Time 1-64")
p5
```

### Filtering outlier from group 2 and Figure 6
```{r}
RATSL8S1 <- RATSL8S %>%
  filter(mean < 550)
  
glimpse(RATSL8S1)

p1 <- ggplot(RATSL8S1, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight), Time 1-64")
p5
```


### Anova model
```{r}
fit <- lm(mean ~ Group, data = RATSL8S1)
summary(fit)
anova(fit)
```



## Analysis of BPRSL dataset

### Reading the data
```{r}
read.csv("BPRSL.csv")
glimpse(BPRSL)
head(BPRSL); tail(BPRSL)
```
### Figure 1

```{r}
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ treatment, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
p6
```

### Figure 2
```{r, echo=TRUE, fig.width=10, fig.height=10}
pairs(BPRS, cex = 0.7)
```
### Anova
```{r, echo=TRUE, fig.width=10, fig.height=10}
library("lme4")
BPRSL_ref <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRSL_ref)
```



```{r, echo=TRUE}
BPRSL_ref1 <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRSL_ref1)
```

```{r, echo=TRUE}
BPRSL_ref2 <- lm(bprs ~ week * treatment, data = BPRSL)
summary(BPRSL_ref2)
```


```{r, echo=TRUE}
anova(BPRSL_ref1, BPRSL_ref2)
```

### Figure 3
```{r, echo=TRUE, fig.width=3, fig.height=4}
library(ggplot2)

Fitted <- fitted(BPRSL_ref2)
BPRSL <- BPRSL %>% mutate(Fitted)
```

```{r}
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, group = treatment))
p2 <- p1 + geom_line(aes(linetype = treatment))
p3 <- p2 + scale_x_continuous(name = "weeks")
p4 <- p3 + scale_y_continuous(name = "bprs")
p5 <- p4 + theme_bw() + theme(legend.position = "right")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p7 <- p6 + ggtitle("Observed")
graph1 <- p7
```


```{r, echo=TRUE, fig.width=3, fig.height=4}
p1 <- ggplot(BPRSL, aes(x = week, y = Fitted, group = treatment))
p2 <- p1 + geom_line(aes(linetype = treatment))
p3 <- p2 + scale_x_continuous(name = "weeks")
p4 <- p3 + scale_y_continuous(name = "bprs")
p5 <- p4 + theme_bw() + theme(legend.position = "right")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p7 <- p6 + ggtitle("Fitted")
graph2 <- p7

graph1; graph2

```

## The Random Intercept Model

library(lme4)
```{r}
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
```

### Summary and Student t-test
```{r}
BPRSL8S <- BPRSL %>%
  filter(week > 0) %>%
  group_by(treatment, subject) %>%
  summarise( mean=mean(bprs) ) %>%
  ungroup()
t.test(mean ~ treatment, data = BPRSL8S, var.equal = TRUE)
```

### Anova

```{r}
baseline <- BPRS$week0
BPRSL8S1 <- BPRSL8S %>%
  mutate(baseline)
  fit <- lm(mean ~ baseline + treatment, data = BPRSL8S1)
summary(fit)
anova(fit)
```
