# Regression and model validation

## Reading and exploring structure and dimension of the data
"In this excercise I first read the data (learning2014.csv) that was created in the data wrankling part of the excercise. For this I used **read.csv()**-commmand that reads the data. To explore structure and dimension of the data, I used following **dim(learning2014)** **str(learning2014)** commands. Dimension command showed that data consisted of 166 observations and seven variables. Stucture command showed the structure of the data. Seven variables were *Gender, Age, Attitude, Seep, Stra, Surf, Points*. Gender consist of two level factor (F=1 or M=2). Age is showed as an years. Attitude, deep, stra and surf are numeric factors. Points -variable indicates the exam points of the person"

##Graphical overview of the data
"For generating graphical overview about the data, **pairs()**-command was used first. In the following graph I used  fast plotting function that draw all possible scatter plots from the column of a data frame. However, gender was excluded from the plots but it was added there with different colours (M=black, F=red)"  

```{r}
pairs(learning2014 [-1], col = learning2014$gender)
```
"This plot is however a bit hard to understand and therefore I made gg-plot about the data:"

```{r}
p <- ggpairs(learning2014, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))
```